openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2023-06-02 2023-05-revision-1/response-to-reviewers.tex

\vspace{10mm}

\subsection*{Reviewer B}

\reviewc{r1q1}{I just have a few minor comments regarding the text. Firstly, the authors claim that FIt-SNE is not a PyPI/conda package in table 1, and that is just not the case. There is a similar comment in the introduction at the bottom of page 2 that needs to be remedied in this regard. "However, FIt-SNE is difficult to install and distribute, and provides and API largely inconsistent with the Python data science ecosystem".  I also find it fairly the comment fairly vague in the context of being inconsistent with Python data-science ecosystem, as the functional approach with most of the parameters being set by default is also a commonly used API for software.}

\answer{
We apologize for this oversight and would like to thank the reviewer for pointing out the availability of PyPI/conda packages for FIt-SNE. We have updated the text and the respective content of Table 1.
}

@@

\reviewc{r1q2}{Finally, this is one of the only works that switches between the Barnes-Hut approach to the FIt-SNE algorithm and the authors mention that they have a heuristic for doing so. This is an extremely useful feature as well, as it chooses the best of both worlds, and it would be wonderful if the authors could provide a brief description for the heuristic they use in their code. }

\answer{We have now included a sentence in Section 3 on page 8 that describes our heuristics.}

@@

\reviewc{r2q2}{There are no details given about preprocessing transcriptomic data (normalization, log-transformation, gene selection, PCA, etc.) As the paper uses RNAseq datasets for all examples, these details should be provided somewhere.}

\answer{We have added the details of the single-cell RNA-seq preprocessing pipeline to the Appendix.
}

\reviewc{r2q3}{page 7: "one can speed up the overall computation by choosing a uniform kernel" -- this sounds cryptic as the uniform kernel was not explained anywhere. It could make sense to elaborate on it somewhere: what exactly is it? how to use it?}

\answer{
We have reworded the paragraph to describe the uniform affinity kernel.
}

@@

\reviewc{r2q4}{page 7: initialization paragraph only mentions PCA, would it make sense to mention that openTSNE also supports Laplacian eigenmaps as initialization?}

\answer{
We now mention this option in the corresponding paragraph on page 8 and have also added a reference to Laplacian eigenmaps in Section 2.4 in the paragraph on global initialization.
}

\reviewc{r2q5}{It would be interesting to add UMAP to the runtime benchmark. UMAP is a very popular alternative to t-SNE, and is often thought to be much faster, which is not actually the case for openTNSE. Perhaps a separate benchmark figure could focus on openTSNE vs UMAP (depending on the sample size)? This may depend on the used affinity model, as uniform affinity with 15 neighbors should be much faster than the standard affinity model (with 90 neighbors) for large sample sizes. If so, it may be of interest to show both and discuss.}

\answer{
Initially, we decided to omit a comparison to UMAP, as it is a different algorithm, and a one-to-one comparison is difficult and not the focal point of the manuscript. However, given the popularity of both algorithms, it makes sense to include a benchmark comparison between their implementations. Given the often misunderstood differences between the two algorithms, it is necessary to provide a throughout discussion of their benchmark results. We have added a subsection in Section 5.2, comparing the performance of the \texttt{openTSNE} and \texttt{umap-learn} libraries. The uniform affinity model is also an interesting and valuable addition to the openTSNE framework, and we have added a comparison to that as well, as it could also be used in real-world scenarios when embedding massive data sets.
}

@@

\reviewc{r2q6}{top of page 5: you talk about attractive forces but they have not been introduced yet. It may make sense to move the equation from secton 2.2 into 2.1 after Eq (5).}

\answer{We would like to thank the reviewer for the suggestion. We have now moved the equation from section 2.2 to the end of section 2.1, and updated the text accordingly.}

@@

\reviewc{r2q7}{page 6: "Special care must be taken to tune the learning rate during optimization" -- the details are missing here. So what values are used in openTSNE?}

\answer{We have expanded the paragraph to elaborate on the instability and what parameter values we use in openTSNE.}

\reviewc{r2q8}{page 13: it is unclear what embedding.prepare\_partial() does here -- how are the new samples initialized, before you are running optimize()? Same for the prepare\_partial call on page 17.}

\answer{We agree and have added a sentence below the code snippet describing what each of the function calls does.}

### 2023-06-03 2023-05-revision-1/main.tex
minor changes in appendix
paper: minor typos

We use the Kullback-Leibler (KL) divergence to measure the agreement
between the two distributions $\mathbf{P}$ and $\mathbf{Q}$
\begin{equation}
C = \text{KL}(\mathbf{P} \mid \mid \mathbf{Q}) = \sum_{ij} p_{ij} \log \frac{p_{ij}}{q_{ij}}.
\label{eq:kl_divergence}
\end{equation}
The objective is to find embeddings $\mathbf{Y}$ that minimize the KL divergence. The corresponding gradient takes the form
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) w_{ij},
\label{eq:tsne_gradient}
\end{equation}
where $w_{ij} = \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}$
and represents the unnormalized $q_{ij}$.
We can rewrite this gradient to
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
\Bigg ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}$. With this formulation, we can interpret t-SNE algorithm as a force-directed layout algorithm, where individual data points act as particles exerting both attractive and repulsive on each other until a state of equilibrium is reached.

@@

Alternatively, empirical evidence indicates that replacing the standard perplexity-based Gaussian affinity model in t-SNE with a uniform affinity model produces markedly similar embeddings to that of standard t-SNE. For instance, running standard t-SNE with the perplexity of $u=30$ constructs a KNNG with $3 \cdot u = 90$ neighbors. However, when using the equivalent uniform affinity model, only $u$ neighbors are needed, making the uniform affinity model attractive for visualizing large data sets.

@@

We here describe the preprocessing steps applied to the single-cell RNA-seq data sets used in Fig.~\ref{fig:macosko}~\citep{macosko2015highly}, Fig.~\ref{fig:tasic}~\citep{tasic2018shared}, Fig.~\ref{fig:cao}~\citep{cao2019single}, and Fig.~\ref{fig:transform}~\citep{hochgerner2018conserved}. The standard single-cell preprocessing pipeline consists of gene filtering, gene selection, data normalization, and dimensionality reduction.

>>>>

We use the Kullback-Leibler (KL) divergence to measure the agreement
between the two distributions $\mathbf{P}$ and $\mathbf{Q}$
\begin{equation}
C = \text{KL}(\mathbf{P} \mid \mid \mathbf{Q}) = \sum_{ij} p_{ij} \log \frac{p_{ij}}{q_{ij}}.
\label{eq:kl_divergence}
\end{equation}
The objective is to find embeddings $\mathbf{Y}$ that minimize the KL divergence. The corresponding gradient takes the form
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) w_{ij},
\label{eq:tsne_gradient}
\end{equation}
where $w_{ij} = \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}$
and represents the unnormalized $q_{ij}$.
We can rewrite this gradient as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
\Bigg ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}$. With this formulation, we can interpret t-SNE algorithm as a force-directed layout algorithm, where individual data points act as particles exerting both attractive and repulsive on each other until a state of equilibrium is reached.

@@

Alternatively, empirical evidence indicates that replacing the standard perplexity-based Gaussian affinity model in t-SNE with a uniform affinity model produces markedly similar embeddings to that of standard t-SNE. For instance, running standard t-SNE with a perplexity of $u=30$ constructs a KNNG with $3 \cdot u = 90$ neighbors. However, when using the equivalent uniform affinity model, only $u$ neighbors are needed, making the uniform affinity model attractive for visualizing large data sets.

@@

We here describe the preprocessing steps applied to the single-cell RNA-seq data sets used in Fig.~\ref{fig:macosko}~\citep{macosko2015highly}, Fig.~\ref{fig:tasic}~\citep{tasic2018shared}, Fig.~\ref{fig:cao}~\citep{cao2019single}, and Fig.~\ref{fig:transform}~\citep{hochgerner2018conserved}. The standard single-cell preprocessing pipeline that have used consists of gene filtering, gene selection, data normalization, and dimensionality reduction.
